Global Distribution of Unintended Pregnancies and Contraceptive Usage

### 2024-02-23 code/eda-code/eda.qmd
Part 2 Update
Updated the plot descriptions in the manuscript. Added additional exploratory data analysis conclusions. Cleaned up the template to include what is relevant to our project and updated the read me files.

```{r}
#load needed packages. make sure they are installed.
library(here) #for data loading/saving
library(dplyr)
library(skimr)
library(ggplot2)
```

@@

```{r}
# skim the data to create summary file
summary_df = skim(mydata)
print(summary_df)
# save to file
summarytable_file = here("results", "tables", "summarytable.rds")
saveRDS(summary_df, file = summarytable_file)
```
We also created simple tables for the categorical variables: `region`, `larcster_usage`, `sarc_usage`, and `none_usage`.
```{r}

```

@@

```{r}
# explore the percentage of unintended pregnancies related to long acting reversible methods in each country
p2 <- mydata %>% ggplot(aes(x=pct_upreg_larcster)) + geom_histogram() 
plot(p2)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_larcster_distribution.png")
ggsave(filename = figure_file, plot=p2)

# explore the percentage of unintended pregnancies related to short acting reversible methods in each country
p3 <- mydata %>% ggplot(aes(x=pct_upreg_sarc)) + geom_histogram() 
plot(p3)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_sarc_distribution.png")
ggsave(filename = figure_file, plot=p3)

# explore the percentage of unintended pregnancies related to traditional methods in each country
p4 <- mydata %>% ggplot(aes(x=pct_upreg_trad)) + geom_histogram() 
plot(p4)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_trad_distribution.png")
ggsave(filename = figure_file, plot=p4)

# explore the percentage of unintended pregnancies related to no birth control usage in each country
p5 <- mydata %>% ggplot(aes(x=pct_upreg_nouse)) + geom_histogram() 
plot(p5)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_nouse_distribution.png")
ggsave(filename = figure_file, plot=p5) 
```

@@

```{r}
# explore the total number of people using long acting reversible methods in each country
p6 <- mydata %>% ggplot(aes(x=total_larcster)) + geom_histogram() 
plot(p6)
figure_file = here("results", "figures", "exploratory-figures", "total_larcster_distribution.png")
ggsave(filename = figure_file, plot=p6)

# explore the total number of people using short acting reversible methods in each country
p7 <- mydata %>% ggplot(aes(x=total_sarc)) + geom_histogram() 
plot(p7)
figure_file = here("results", "figures", "exploratory-figures", "total_sarc_distribution.png")
ggsave(filename = figure_file, plot=p7)

# explore the total number of people using traditional methods in each country
p8 <- mydata %>% ggplot(aes(x=total_trad)) + geom_histogram() 
plot(p8)
figure_file = here("results", "figures", "exploratory-figures", "total_trad_distribution.png")
ggsave(filename = figure_file, plot=p8)

# explore the total number of people not using birth control in each country
p9 <- mydata %>% ggplot(aes(x=total_none)) + geom_histogram() 
plot(p9)
figure_file = here("results", "figures", "exploratory-figures", "total_none_distribution.png")
ggsave(filename = figure_file, plot=p9) 
```

We need to explore where the outliers in these variables are coming from. China and India are the two countries that account for the outliers in all 4 variables, which is likely being skewed by population size. The first method of standardizing we tried is creating proportions using the total count for each birth control method type. After creating these standardized variables, the same issue with outliers remained because China and India still account for the largest proportion of observations. Next, we considered using a scaling factor. We attempted this approach with the `total_larcster` variable and found the same issues as before. The third attempt at standarization utilized a min-max scaling with z-score normalization, but the outlier issue remained. 
```{r}
# find the countries that represent the outliers
outliers <- mydata %>% filter(total_larcster > 25000, 
                              total_sarc > 25000,
                              total_trad > 5000,
                              total_none > 10000)

@@

We will start exploring how many women use each type of birth control in each country compared to the percent of unintended pregnancies related to that birth control. Plot 10 shows `pct_upreg_larcster` vs `larcster_standard` to analyze women using long acting reversible methods and sterilization versus the percentage of unintended pregnancies using those methods. Plot 11 uses `pct_upreg_sarc` and `sarc_standard` to examine the same relationship as plot 10 but with short-acting reversible methods. Plot 12 shows `pct_upreg_trad` versus `trad_standard`, and plot 13 shows `pct_upreg_nouse` versus `none_standard`. 
```{r}
# explore percentage of unintended pregnancies compared to standardized count of women using long-acting reversible and sterilization methods
p10 <- mydata %>% ggplot(aes(x=pct_upreg_larcster, y=larcster_standard)) + geom_point() + geom_smooth(method='lm')
plot(p10)
figure_file = here("results", "figures", "exploratory-figures", "pctuppreg_totallarcster.png")
ggsave(filename = figure_file, plot=p10)

# explore percentage of unintended pregnancies compared to standardized count of women using short acting reversible methods
p11 <- mydata %>% ggplot(aes(x=pct_upreg_sarc, y=sarc_standard)) + geom_point() + geom_smooth(method='lm')
plot(p11)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_totalsarc.png")
ggsave(filename = figure_file, plot=p11)

# explore percentage of unintended pregnancies compared to standardized count of women using traditional methods
p12 <- mydata %>% ggplot(aes(x=pct_upreg_trad, y=trad_standard)) + geom_point() + geom_smooth(method='lm')
plot(p12)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_totaltrad.png")
ggsave(filename = figure_file, plot=p12)

# explore percentage of unintended pregnancies compared to standardized count of women not using no method
p13 <- mydata %>% ggplot(aes(x=pct_upreg_nouse, y=none_standard)) + geom_point() + geom_smooth(method='lm')
plot(p13)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_totalnone.png")
ggsave(filename = figure_file, plot=p13) 
```

@@

```{r}
# explore percent unintended pregnancies using long acting reversible and sterilziation methods by region
p14 <- mydata %>% ggplot(aes(x=pct_upreg_larcster, y=larcster_standard, color = region)) + geom_point() + geom_smooth(method='lm')
plot(p14)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_larcster_stratified.png")
ggsave(filename = figure_file, plot=p14)

# explore percent unintended pregnancies using short acting reversible methods by region
p15 <- mydata %>% ggplot(aes(x=pct_upreg_sarc, y=sarc_standard, color = region)) + geom_point() + geom_smooth(method='lm')
plot(p15)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_sarc_stratified.png")
ggsave(filename = figure_file, plot=p15)

# explore percent unintended pregnancies using traditional methods by region
p16 <- mydata %>% ggplot(aes(x=pct_upreg_trad, y=trad_standard, color = region)) + geom_point() + geom_smooth(method='lm')
plot(p16)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_trad_stratified.png")
ggsave(filename = figure_file, plot=p16)

# explore percent unintended pregnancies using no method by region
p17 <- mydata %>% ggplot(aes(x=pct_upreg_nouse, y=none_standard, color = region)) + geom_point() + geom_smooth(method='lm')
plot(p17)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_none_stratified.png")
ggsave(filename = figure_file, plot=p17) 
```

Our next group of plots will be violin plots to explore the usage level of each birth control method, a categorical variable, associated with percentage of unintended pregnancies use for each birth control method, a numerical variable.
```{r}
# explore long acting reversible and sterilization method usage compared to percent of unintended pregnancies
p18 <- mydata %>% ggplot(aes(x = larcster_usage, y = pct_upreg_larcster)) +
  geom_violin() +
  labs(x = "Long Acting Reversible and Sterilization Usage", y = "Percentage Unintended Pregnancies") 
plot(p18)
figure_file = here("results", "figures", "exploratory-figures", "larcster_pctupreg_violin.png")
ggsave(filename = figure_file, plot=p18)

# explore short acting reversible method usage compared to percent of unintended pregnancies
p19 <- mydata %>% ggplot(aes(x = sarc_usage, y = pct_upreg_sarc)) +
  geom_violin() +
  labs(x = "Short Acting Reversible Usage", y = "Percentage Unintended Pregnancies") 
plot(p19)
figure_file = here("results", "figures", "exploratory-figures", "sarc_pctupreg_violin.png")
ggsave(filename = figure_file, plot=p19)

# explore traditional method usage compared to percent of unintended pregnancies
p20 <- mydata %>% ggplot(aes(x = trad_usage, y = pct_upreg_trad)) +
  geom_violin() +
  labs(x = "Traditional Method Usage", y = "Percentage Unintended Pregnancies") 
plot(p20)
figure_file = here("results", "figures", "exploratory-figures", "trad_pctupreg_violin.png")
ggsave(filename = figure_file, plot=p20)

# explore no method usage compared to percent of unintended pregnancies
p21 <- mydata %>% ggplot(aes(x = none_usage, y = pct_upreg_nouse)) +
  geom_violin() +
  labs(x = "No Birth Control Usage", y = "Percentage Unintended Pregnancies") 
plot(p21)
figure_file = here("results", "figures", "exploratory-figures", "none_pctupreg_violin.png")
ggsave(filename = figure_file, plot=p21) 
```

>>>>

```{r}
#load needed packages. make sure they are installed.
library(here) #for data loading/saving
library(dplyr) # for basic piping functions
library(skimr) # for summarizing data
library(ggplot2) # for visualizing data
library(gtsummary) # for creating tables
```

@@

```{r}
# skim the data to create summary file
summary_df <- skim(mydata)
print(summary_df)
# save to file
summarytable_file <- here("results", "tables", "summarytable.rds")
saveRDS(summary_df, file = summarytable_file)
```

We also created a simple table with the categorical variables: `region`, `larcster_usage`, `sarc_usage`, and `none_usage`.
```{r}
# isolate categorical variables into their own data set
mydata_cat <- mydata %>% select(region, larcster_usage, sarc_usage, trad_usage, none_usage)

# create a summary table with counts of each level of each categorical variable
summary_table <- tbl_summary(mydata_cat, statistic = all_categorical() ~ "{n}")
print(summary_table)
```

@@

```{r}
# explore the percentage of unintended pregnancies related to long acting reversible methods in each country
p2 <- mydata %>% ggplot(aes(x=pct_upreg_larcster)) + 
  geom_histogram() +
  ggtitle("Percent Unintended Pregnancy Using Long Acting Reversible Methods") + xlab("Percentage") + ylab("Number of Countries")
plot(p2)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_larcster_distribution.png")
ggsave(filename = figure_file, plot=p2)

# explore the percentage of unintended pregnancies related to short acting reversible methods in each country
p3 <- mydata %>% ggplot(aes(x=pct_upreg_sarc)) + 
  geom_histogram() + 
  ggtitle("Percent Unintended Pregnancy Using Short Acting Reversible Methods") + xlab("Percentage") + ylab("Number of Countries")
plot(p3)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_sarc_distribution.png")
ggsave(filename = figure_file, plot=p3)

# explore the percentage of unintended pregnancies related to traditional methods in each country
p4 <- mydata %>% ggplot(aes(x=pct_upreg_trad)) + 
  geom_histogram() + 
  ggtitle("Percent Unintended Pregnancy Using Traditional Methods") + xlab("Percentage") + ylab("Number of Countries")
plot(p4)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_trad_distribution.png")
ggsave(filename = figure_file, plot=p4)

# explore the percentage of unintended pregnancies related to no birth control usage in each country
p5 <- mydata %>% ggplot(aes(x=pct_upreg_nouse)) +
  geom_histogram() + 
  ggtitle("Percent Unintended Pregnancy Using No Birth Control") + xlab("Percentage") + ylab("Number of countries")
plot(p5)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_nouse_distribution.png")
ggsave(filename = figure_file, plot=p5) 
```
P2: This histogram is highly skewed right, which shows that most countries experience a very low percentage of unintended pregnancies related to long acting reversible and sterilization methods.

P3: This histogram shows that the percentage of unintended pregnancies among those using short acting reversible methods varies more between countries compared to long acting reversible methods. About half of the countries in the dataset report less than 20% of pregnancies are unintended when women use this form of birth control.

P4: This histogram shows that about half of the countries reported less than 10% percent of unintended pregnancies associated with using traditional birth control methods

P5: This histogram shows that more countries report a higher percentage of unintended pregnancies among women who do not use any contraception, but it varies widely by country.

@@

```{r}
# explore the total number of people using long acting reversible methods in each country
p6 <- mydata %>% ggplot(aes(x=total_larcster)) + 
  geom_histogram() + 
  ggtitle("Total Usage of Long Acting Reversible Methods per Country") + xlab("Number of Women") + ylab("Number of Countries")
plot(p6)
figure_file = here("results", "figures", "exploratory-figures", "total_larcster_distribution.png")
ggsave(filename = figure_file, plot=p6)

# explore the total number of people using short acting reversible methods in each country
p7 <- mydata %>% ggplot(aes(x=total_sarc)) + 
  geom_histogram() + 
  ggtitle("Total Usage of Short Acting Reversible Methods per Country") + xlab("Number of Women") + ylab("Number of Countries")
plot(p7)
figure_file = here("results", "figures", "exploratory-figures", "total_sarc_distribution.png")
ggsave(filename = figure_file, plot=p7)

# explore the total number of people using traditional methods in each country
p8 <- mydata %>% ggplot(aes(x=total_trad)) + 
  geom_histogram() + 
  ggtitle("Total Usage of Traditional Methods per Country") + xlab("Number of Women") + ylab("Number of Countries")
plot(p8)
figure_file = here("results", "figures", "exploratory-figures", "total_trad_distribution.png")
ggsave(filename = figure_file, plot=p8)

# explore the total number of people not using birth control in each country
p9 <- mydata %>% ggplot(aes(x=total_none)) + 
  geom_histogram() + 
  ggtitle("Total Usage of No Birth Control Method per Country") + xlab("Number of Women") + ylab("Number of Countries")
plot(p9)
figure_file = here("results", "figures", "exploratory-figures", "total_none_distribution.png")
ggsave(filename = figure_file, plot=p9) 
```
The two outliers present in each of these graphs makes them very difficult to interpret. We need to explore where the outliers in these variables are coming from. China and India are the two countries that account for the outliers in all 4 variables, which is likely being skewed by population size. The first method of standardizing we tried is creating proportions using the total count for each birth control method type. After creating these standardized variables, the same issue with outliers remained because China and India still account for the largest proportion of observations. Next, we considered using a scaling factor. We attempted this approach with the `total_larcster` variable and found the same issues as before. The third attempt at standarization utilized a min-max scaling with z-score normalization, but the outlier issue remained. 
```{r}
# find the countries that represent the outliers
outliers <- mydata %>% filter(total_larcster > 25000, 
                              total_sarc > 25000,
                              total_trad > 5000,
                              total_none > 10000)

@@

We will start exploring how many women use each type of birth control in each country compared to the percent of unintended pregnancies related to that birth control. Plot 10 shows `pct_upreg_larcster` vs `larcster_standard` to analyze women using long acting reversible methods and sterilization versus the percentage of unintended pregnancies using those methods. Plot 11 uses `pct_upreg_sarc` and `sarc_standard` to examine the same relationship as plot 10 but with short-acting reversible methods. Plot 12 shows `pct_upreg_trad` versus `trad_standard`, and plot 13 shows `pct_upreg_nouse` versus `none_standard`. 
```{r}
# explore percentage of unintended pregnancies compared to standardized count of women using long-acting reversible and sterilization methods
p10 <- mydata %>% ggplot(aes(x=pct_upreg_larcster, y=larcster_standard)) + 
  geom_point() + 
  geom_smooth(method='lm') + 
  ggtitle("Percent of Unintended Pregnancies vs Standardzied Count of Women Using LARCs") + xlab("Percentage of Unintended Pregnancies") + ylab("Standarized Total Number of Women")
plot(p10)
figure_file = here("results", "figures", "exploratory-figures", "pctuppreg_totallarcster.png")
ggsave(filename = figure_file, plot=p10)

# explore percentage of unintended pregnancies compared to standardized count of women using short acting reversible methods
p11 <- mydata %>% ggplot(aes(x=pct_upreg_sarc, y=sarc_standard)) + 
  geom_point() + 
  geom_smooth(method='lm') + 
  ggtitle("Percent of Unintended Pregnancies vs Standardzied Count of Women Using SARCs") + xlab("Percentage of Unintended Pregnancies") + ylab("Standarized Total Number of Women")
plot(p11)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_totalsarc.png")
ggsave(filename = figure_file, plot=p11)

# explore percentage of unintended pregnancies compared to standardized count of women using traditional methods
p12 <- mydata %>% ggplot(aes(x=pct_upreg_trad, y=trad_standard)) + 
  geom_point() + 
  geom_smooth(method='lm') + 
  ggtitle("Percent of Unintended Pregnancies vs Standardzied Count of Women Using Traditional Methods") + xlab("Percentage of Unintended Pregnancies") + ylab("Standarized Total Number of Women")
plot(p12)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_totaltrad.png")
ggsave(filename = figure_file, plot=p12)

# explore percentage of unintended pregnancies compared to standardized count of women not using no method
p13 <- mydata %>% ggplot(aes(x=pct_upreg_nouse, y=none_standard)) + 
  geom_point() + 
  geom_smooth(method='lm') + 
  ggtitle("Percent of Unintended Pregnancies vs Standardzied Count of Women Using No Method") + xlab("Percentage of Unintended Pregnancies") + ylab("Standarized Total Number of Women")
plot(p13)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_totalnone.png")
ggsave(filename = figure_file, plot=p13) 
```

P10: As the standardized count of women using LARCs and sterilization methods increases, the percentage of unintended pregnancies tends to increase. There is minimal variation around the linear equation, excluding the two largest percentages accounted for by India and China.

P11: The same positive trend is shown between the percentage of unintended pregnancies and number of women using SARMs, but there is considerably more variation in this relationship.

P12: As percentage of unintended pregnancies associated with traditional methods increases, the total number of women using this method increases. There is minimal variation in the trend when percentage of unintended pregnancies is less than 8%, but there is more variation when the percentage is greater than 10%.

P13: The same expected positive trend is observed for women who are not using a birth control method, but there is considerable variation in this graph across all percentages of unintended pregnancies. Additionally, the lowest percentage in this category is around 35%, which is much higher than the percentages associated with the three birth control methods.

@@

```{r}
# explore percent unintended pregnancies using long acting reversible and sterilziation methods by region
p14 <- mydata %>% ggplot(aes(x=pct_upreg_larcster, y=larcster_standard, color = region)) + 
  geom_point() + geom_smooth(method='lm') + 
  ggtitle("Percent of Unintended Pregnancies vs Standardzied Count of Women Using LARCs") + xlab("Percentage of Unintended Pregnancies") + ylab("Standarized Total Number of Women")
plot(p14)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_larcster_stratified.png")
ggsave(filename = figure_file, plot=p14)

# explore percent unintended pregnancies using short acting reversible methods by region
p15 <- mydata %>% ggplot(aes(x=pct_upreg_sarc, y=sarc_standard, color = region)) + 
  geom_point() + 
  geom_smooth(method='lm') + 
  ggtitle("Percent of Unintended Pregnancies vs Standardzied Count of Women Using SARCs") + xlab("Percentage of Unintended Pregnancies") + ylab("Standarized Total Number of Women")
plot(p15)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_sarc_stratified.png")
ggsave(filename = figure_file, plot=p15)

# explore percent unintended pregnancies using traditional methods by region
p16 <- mydata %>% ggplot(aes(x=pct_upreg_trad, y=trad_standard, color = region)) + 
  geom_point() + 
  geom_smooth(method='lm') +  
  ggtitle("Percent of Unintended Pregnancies vs Standardzied Count of Women Using Traditional Methods") + xlab("Percentage of Unintended Pregnancies") + ylab("Standarized Total Number of Women")
plot(p16)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_trad_stratified.png")
ggsave(filename = figure_file, plot=p16)

# explore percent unintended pregnancies using no method by region
p17 <- mydata %>% ggplot(aes(x=pct_upreg_nouse, y=none_standard, color = region)) + 
  geom_point() + 
  geom_smooth(method='lm') +   
  ggtitle("Percent of Unintended Pregnancies vs Standardzied Count of Women Using No Method") + xlab("Percentage of Unintended Pregnancies") + ylab("Standarized Total Number of Women")
plot(p17)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_none_stratified.png")
ggsave(filename = figure_file, plot=p17) 
```
P14: Among women using long acting reversible and sterilization methods, Asia has the highest percentages of unintended pregnancies, followed by Latin America/the Caribbean and Africa.

P15: This graph for short acting reversible methods shows clearer differences between regions, and Latin America/the Caribbean leads in the percentage of unintended pregnancies in this category, but Africa shows the sharpest increase in percentage of unintended pregnancies as the total population increases.

P16: Among women using traditional birth control methods, Europe has the sharpest increase in unintended pregnancies as population increases, but Asia has the highest percentage of unintended pregnancies in this category.

P17: This graph is much different than the other three, which is to be expected at this point. Oceania has the highest percentage of unintended pregnancies when women are not using birth control. The slope of the relationship for each region appears to be similar in this graph with the exception of Oceania having a steeper slope. There are considerably more outliers in this graph with African countries

Our next group of plots will be violin plots to explore the usage level of each birth control method, a categorical variable, associated with percentage of unintended pregnancies use for each birth control method, a numerical variable.
```{r}
# explore long acting reversible and sterilization method usage compared to percent of unintended pregnancies
p18 <- mydata %>% ggplot(aes(x = larcster_usage, y = pct_upreg_larcster)) +
  geom_violin() +
  ggtitle("Usage of LARCs and Sterilization vs Percent of Unintended Pregnancies") +
  labs(x = "Long Acting Reversible and Sterilization Usage", y = "Percentage Unintended Pregnancies") 
plot(p18)
figure_file = here("results", "figures", "exploratory-figures", "larcster_pctupreg_violin.png")
ggsave(filename = figure_file, plot=p18)

# explore short acting reversible method usage compared to percent of unintended pregnancies
p19 <- mydata %>% ggplot(aes(x = sarc_usage, y = pct_upreg_sarc)) +
  geom_violin() + ggtitle("Usage of SARCs vs Percent of Unintended Pregnancies") +
  labs(x = "Short Acting Reversible Usage", y = "Percentage Unintended Pregnancies") 
plot(p19)
figure_file = here("results", "figures", "exploratory-figures", "sarc_pctupreg_violin.png")
ggsave(filename = figure_file, plot=p19)

# explore traditional method usage compared to percent of unintended pregnancies
p20 <- mydata %>% ggplot(aes(x = trad_usage, y = pct_upreg_trad)) +
  geom_violin() + 
  ggtitle("Usage of Traditional Method vs Percent of Unintended Pregnancies") +
  labs(x = "Traditional Method Usage", y = "Percentage Unintended Pregnancies") 
plot(p20)
figure_file = here("results", "figures", "exploratory-figures", "trad_pctupreg_violin.png")
ggsave(filename = figure_file, plot=p20)

# explore no method usage compared to percent of unintended pregnancies
p21 <- mydata %>% ggplot(aes(x = none_usage, y = pct_upreg_nouse)) +
  geom_violin() +  
  ggtitle("No Birth Control Usage vs Percent of Unintended Pregnancies") +
  labs(x = "No Birth Control Usage", y = "Percentage Unintended Pregnancies") 
plot(p21)
figure_file = here("results", "figures", "exploratory-figures", "none_pctupreg_violin.png")
ggsave(filename = figure_file, plot=p21) 
```
P18: The distribution of percentage of unintended pregnancies varies a lot among usage levels of long acting reversible and sterilization methods. High usage countries have a consistent percentage of unintended pregnancies, but countries with a medium usage level tend to have lower percentages of unintended pregnancies.

P19: It appears that the distribution of short acting reversible method usage is skewed towards lower percentages of unintended pregnancies across all usage levels, suggesting that more countries have lower percentage of unintended pregnancies among women using this method.

P20: Among women using traditional birth control methods, the distribution of percentage of unintended pregnancies is similar among medium and high usage countries, but low usage countries are skewed towards lower percentage of unintended pregnancies.

P21: While the distribution of percentage of unintended pregnancies is similar across the three usage levels for women not using any birth control method, the high usage countries appear to have a bottleneck in percentage of unintended pregnancies. The high and low usage countries appear to be skewed towards higher levels of unintended pregnancies, but the medium usage countries appear to have lower percentage of unintended pregnancies.

### 2024-02-23 products/manuscript/Manuscript.qmd
Part 2 Update
Updated the plot descriptions in the manuscript. Added additional exploratory data analysis conclusions. Cleaned up the template to include what is relevant to our project and updated the read me files.

This uses MS Word as output format. [See here](https://quarto.org/docs/output-formats/ms-word.html) for more information. You can switch to other formats, like html or pdf. See [the Quarto documentation](https://quarto.org/) for other formats.

@@

@@

To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above (here `dataanalysis_template_references.bib`) and have the right bibtex key. Then you can include like this:

Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].

>>>>

This uses MS Word as output format. [See here](https://quarto.org/docs/output-formats/ms-word.html) for more information.

@@

# Project Part 2
# Exploring and processing the data 
The first priority when cleaning the data was analyzing missing values. We found that the `pct_upreg` variable containing the percentage of unintended pregnancies is unreliable due to missing data that could not be explained. We decided to use the variables with percentage of unintended pregnancies related to each birth control method as our outcome variables because they are complete and were found directly based on the AIU protocol information. We printed the summary table here becacuse it was useful in determining missingness and creating a new factor variable for the usage of each birth control method at the factor level later in the processing sequence. 
```{r}
# show summary table to numeric variables with misssing values
print(summary_df)
```
We explored the distribution of the countries by five categorical variables: `region`, `larcster_usage`, `sarc_usage`, `trad_usage`, and `none_usage`. We found that most countries are located in Africa followed by Asia and Latin America/the Caribbean. Only 20 countries have high usage of long acting reversible and sterilization methods compared to 33 countries with high usage of short acting reversible and traditional methods. It is interesting that the distributions for short acting reversible method, traditional method, and no method usages are the same. 
```{r}
print(summary_table)
```

We created a multitude of exploratory plots to understand our data better. The histograms of percent of unintended pregnancies related to each method of birth control showed discrepancies in skewness. The `pct_upreg_larcster` and `pct_upreg_trad` variables are highly skewed right, which shows that more countries experience lower rates of unintended pregnancies when using these methods. The `pct_upreg_sarc` histogram is not as skewed. It shows that the percentage of unintended pregnancies among those using short acting reversible methods varies more between countries compared to long acting reversible methods. The histogram for `pct_upreg_none` shows that more countries report a higher percentage of unintended pregnancies among women who do not use any contraception, but it varies widely by country. 
```{r}
plot(p2)
plot(p3)
plot(p4)
plot(p5)
```

The next set of variables we explored were the total counts of women using each birth control method. There were two large outliers that prevented the graphs from being useful. After exploring the outliers, we standardized using the population size of reproductive aged women (15-49 years). There is a positive association between the percentage of unintended pregnancies and the total number of women using that method of birth control for all three methods and those not using any method, which is to be expected. There is more variation among those using SARMs and traditional methods compared to those using long-acting reversible methods. Additionally, the lowest percentage in this category is around 35%, which is much higher than the percentages associated with the three birth control methods. 
```{r}
plot(p10)
plot(p11)
plot(p12)
plot(p13)
```
We stratified these four graphs by region to examine how the relationships differ across each of the 6 regions. The long acting reversible and sterilization methods graph is a little difficult to decipher, but Asia has the highest percentages of unintended pregnancies, followed by Latin America/the Caribbean and Africa. The graph for short acting reversible methods shows clearer differences between regions, and Latin America/the Caribbean leads in the percentage of unintended pregnancies in this category, but Africa shows the sharpest increase in percentage of unintended pregnancies as the total population increases.Among women using traditional birth control methods, Europe has the sharpest increase in unintended pregnancies as population increases, but Asia has the highest percentage of unintended pregnancies in this category. The graph for women not using any birth control methods shows that Oceania has the highest percentage of unintended pregnancies. 
```{r}
plot(p14)
plot(p15)
plot(p16)
plot(p17)
```
The last set of plots we created to explore the data are violin plots using the categorical usage variables for each birth control method. The first violin plot shows that the distribution of percentage of unintended pregnancies varies a lot among usage levels of long acting reversible and sterilization methods. High usage countries have a consistent percentage of unintended pregnancies, but countries with a medium usage level tend to have lower percentages of unintended pregnancies. It appears that the distribution of short acting reversible method usage is skewed towards lower percentages of unintended pregnancies across all usage levels, suggesting that more countries have lower percentage of unintended pregnancies among women using this method. Among women using traditional birth control methods, the distribution of percentage of unintended pregnancies is similar among medium and high usage countries, but low usage countries are skewed towards lower percentage of unintended pregnancies. While the distribution of percentage of unintended pregnancies is similar across the three usage levels for women not using any birth control method, the high usage countries appear to have a bottleneck in percentage of unintended pregnancies. The high and low usage countries appear to be skewed towards higher levels of unintended pregnancies, but the medium usage countries appear to have lower percentage of unintended pregnancies. 
```{r}
plot(p18)
plot(p19)
plot(p20)
plot(p21)
```

@@

To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above (here `dataanalysis_template_references.bib`) and have the right bibtex key. Then you can include like this:
Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].
